---
title: Matrixdist News
author: gzt
date: '2018-02-16'
slug: matrixdist-new
categories:
  - R
tags:
  - R
  - matrixdist
---

I have just finished setting up handling for AR(1) and compound symmetry variance structures in my maximum likelihood estimation function for matrix variate normal distributions in `matrixdist`. This means I may submit it to CRAN soon (it's currently available on [github](http://www.github.com/gzt/matrixdist)).  An example:

```{r matrixexample, cache=T}
library(matrixdist)
A <- rmatrixnorm(100, mean=array(0,dim=c(3,4)), U = toeplitz(c(1,.8,.64)), V = rWishart(1,7,diag(4))[,,1])
mle.matrixnorm(A, row.variance="AR(1)")

```

There is a performance hit as fitting for these require calls to `uniroot` to find the zeros of the derivatives, but it is not so bad in the cases I've looked at.

```{r benchmarking, cache=TRUE}
library(microbenchmark)
res<-microbenchmark(
  mle.matrixnorm(A),
  mle.matrixnorm(A, row.variance="AR(1)"),
  mle.matrixnorm(A, row.variance="CS")
)
plot(res, log="y", names = c("None", "AR(1)", "CS"), 
     col = c("darkred","royalblue","lightyellow"),
     xlab="Variance Structure")
```

I will need to think hard about whether I want to add anything else first, as nobody wants to deal with massive updates to functionality in fast succession.

Here is a possible agenda:
 * More distributions, such as skew $t$-distributions. I do not have much interest in matrix variate gamma distributions or beta distributions.
 * Tools for working with matrix variate distributions, such as adaptations for LDA-type activities or clustering.
 * Parameter estimation for distributions besides the normal distribution. $t$-distributions typically require use of EM or its extensions.
